Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-09 paper/paper_draft.qmd
Updates to incorporate tracked changes and add in ADRIA-CoralBlox information

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of note is that the representation of bleaching
mortalities followed Bozec et al., [-@bozec2022] and albeit with a scaling coefficient to
incorporate the amelioration effect that depth may have, following Baird et al.,
[-@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high
heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of
Thorns Starfish or coral disease were not included in this investigation in order to focus
on assessing reef resilience under possible future warming scenarios.

@@

Individual feature analysis for carrying capacity indicates that its influence on cluster
assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This
aligns with the right-skewed distribution of carrying capacity across reefs for these three
predictor variables. Beyond this low-end range, partial dependence becomes near constant,
indicating outlier values for these factors are not influential. Assessing random forest
prediction accuracy across bioregions, reveals increasing performance as mean bioregion reef
depth increases (@sfig-rf-predictor-perf). Unexplained variation in the test dataset (33% of
variation) may be concentrated in shallow bioregions where considered model predictors do
not capture differences between clustered reefs.

@@

Spatially, these results indicate that regions of positive carbonate budget maintenance,
under ACCESS-ESM1-5, and high depth levels were concentrated in the Far Northern, offshore
Townsville/Whitsunday, and Mackay/Capricorn management areas (@sfig-depth-map;
@sfig-carb-budget-map). Resilient reefs in the Far Northern management area were exposed to
lower mean DHW levels than other reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however
similar spatial patterns were observed across GCMs for reefs clustered into the high cover
cluster (@sfig-gcm-cluster-map).

@@

The CoralBlox model incorporates linear extension (growth) and background mortality rates
that are dependent on the size and functional group of a coral colony. Environmental
disturbances and reproduction are not directly considered within CoralBlox. Instead, these
factors are considered by making changes to the distribution of colonies within CoralBlox
timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform
[@iwanaga2023]. Within ADRIA, we incorporate coral life history and environmental events in
the following order: coral colony growth; coral reproduction; larval connectivity and
settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled
together to represent the system, we refer to both as ADRIA-CoralBlox.

Coral fecundity rates are dependent on colony size and functional group following the
results of Hall & Hughes [-@hall1996]. Coral fertilization rates are not considered within
this model, however larval survival and viability are incorporated in connectivity modelling
(see Connectivity modelling). Viable larvae settle at a reef location and are incorporated
into the model as the smallest size class but are not directly impacted by modelled
bleaching mortality. There has been contrasting evidence for juvenile corals being immune to
thermal bleaching events [@depczynski2013], we assume the high background mortality of
juveniles includes thermal stress [@álvarez-noriega2018].

@@

### Model parameterisation

In this investigation we model five functional groups common to the GBR, tabular Acropora,
corymbose Acropora, corymbose non-Acropora, small massives and large massives. We do not
model arborescent Acropora forms due to their distinct life history from other morphological
groups. CoralBlox uses user-defined size classes that influence size-dependent growth and
mortality rates, through which the coral colonies progress as they grow in diameter. Within
ADRIA, we define the distributions of these parameters, and the size classes based on
ecological monitoring data from the Ecological Intelligence for Reef Restoration Program
(EcoRRAP) ([@australianinstituteofmarinescience2025; @ribeirodealmeida2024]).

In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines the proportion of the population on a reef that is
susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

### Calibration

The growth and mortality rate parameters of ADRIA-CoralBlox were calibrated using hard coral
cover and benthic composition data from the Long-Term Monitoring Program (LTMP). From all
the reefs for which there is LTMP data available, 99 were selected for having at least 4
observations that span more than 10 years. These locations were randomly split into a
calibration set (with 73 reefs) and a validation set (with 26 reefs), as evenly spatially
distributed as possible. The error function used in the calibration process is composed of
three elements [@eq-calibration]. The first is a Mean Absolute Exponential Error ($MAEE$)
that captures the difference between modeled and observed hard coral cover. The second is a
Coral Composition Error ($CCE$) that captures the differences in the benthic composition
between modeled and observed coral cover. The third ($Err~p~$) is an error that emphasizes
differences between observed and modelled cover at the beginning and end of the series, as
well as on peaks and troughs:

$$
err = MAEE_\sigma_  *  w_1_  +  CEE_t_  *  w_c_  +  Err_p_  *  w_p_
$$ {#eq-calibration}

[@sfig-reef-scale-calibration] shows a comparison between modeled and observed relative
coral cover for 6 reefs with distinct disturbance patterns. For each one, a Root Mean Square
Error (RMSE), the MAEE and Spearman’s Rank Correlation Coefficient (SRCC) is also shown.
[@sfig-man-area-calibration] presents another comparison between observed and modeled hard
coral cover aggregated by GBRMPA management area.

![Time series comparing modelled and observed relative coral cover for selected validation
reefs. These reefs were selected to show distinct disturbance patterns and behaviors. reefs
(a) and (b), with low heat stress and a single disturbance at the beginning, show the
model’s ability of reproducing the reef’s recovery under low or no disturbance; reefs (c)
and (d), with moderate to high heat stress and disturbance events in the middle of the
considered timeframe, show the model’s ability to reproduce disturbances reasonably well but
some difficulty in representing the post-disturbance recovery ; reefs (e) and (f) illustrate
the challenge of reproducing the observed behavior in a high disturbance
scenario.](../figs/reef_scale_calibration.png){#sfig-reef-scale-calibration
fig-align="center" width="14.82cm"}

![Comparison between observed and modelled coral relative cover aggregated, for each GBRMPA
management area, for the calibration set (first row) and validation set (second row). For
each one the aggregated RMSE and MAE are
presented.](../figs/management_area_calibration.png){#sfig-man-area-calibration
fig-align="center" width="14.82cm"}

>>>>

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of note is that the representation of bleaching
mortalities followed Bozec et al., [-@bozec2022] and Baird et al. [-@baird2018] which
applies a scaling coefficient to incorporate the amelioration effect that depth may have.
This depth-scaling reduces the proportion of coral colonies impacted by high heat stress as
depth increases. The impacts of tropical cyclones, wave activity, Crown of Thorns Starfish
or coral disease were not included in this investigation in order to focus on assessing reef
resilience under possible future warming scenarios.

@@

Individual feature analysis for carrying capacity indicates that its influence on cluster
assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This
aligns with the right-skewed distribution of carrying capacity across reefs for these three
predictor variables. Beyond this low-end range, partial dependence becomes near constant,
indicating outlier values for these factors are not influential. Assessing random forest
prediction accuracy across bioregions reveals increasing performance as mean bioregion reef
depth increases (@sfig-rf-predictor-perf). Unexplained variation in the test dataset (33% of
variation) may be concentrated in shallow bioregions where considered model predictors do
not capture differences between clustered reefs.

@@

Spatially these results indicate that regions of positive carbonate budget maintenance,
under ACCESS-ESM1-5, and high depth levels were concentrated in the Far Northern, offshore
Townsville/Whitsunday, and Mackay/Capricorn management areas (@sfig-depth-map;
@sfig-carb-budget-map). Resilient reefs in the Far Northern management area were exposed to
lower mean DHW levels than other reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however
similar spatial patterns were observed across GCMs for reefs clustered into the high cover
cluster (@sfig-gcm-cluster-map).

@@

The CoralBlox model incorporates linear extension (growth) and background mortality rates
that are dependent on the size and functional group of a coral colony. Environmental
disturbances and reproduction are not directly considered within CoralBlox, and are instead
provided by the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform
[@iwanaga2023]. Within ADRIA, coral life history and environmental events are implemented in
the following order: coral colony growth; coral reproduction; larval connectivity and
settlement; and coral bleaching due to heat stress.

Coral fecundity rates are dependent on colony size and functional group following the
results of Hall & Hughes [-@hall1996]. Coral fertilization rates are not considered within
this model, however larval survival and viability are incorporated in connectivity modelling
(see Connectivity modelling). Viable larvae settle at a reef location and are incorporated
into the model as the smallest size class but are not directly impacted by modelled
bleaching mortality. There has been contrasting evidence for juvenile corals being immune to
thermal bleaching events [@depczynski2013], it is assumed the high background mortality of
juveniles includes thermal stress [@álvarez-noriega2018].

### Model implementation

ADRIA-CoralBlox implements five functional groups common to the GBR, tabular Acropora,
corymbose Acropora, corymbose non-Acropora, small massives and large massives. CoralBlox
uses user-defined size classes that influence size-dependent growth and mortality rates,
through which the coral colonies progress as they grow in diameter. Size classes were
initially informed by Bozec et al. [-@bozec2022].

The area of each coral colony is tracked in CoralBlox by the area of a circumference. A
coral population is then represented by a discretised distribution of **eq** coral
diameters, composed of smaller uniform distributions, each representing the number of corals
within a certain diameter range:

**eq**

where **eq** represents the number of intervals used to discretise **eq**, and **eq** and
**eq** are the diameters of the smallest and biggest coral colonies, in that order. The
coral cover for the entire population is given by:

**eq**

These are uniform distributions that compose **eq**, a coral block characteristed by a
diameter density **eq** (the block's height) and a diameter range **eq** (the base of the
block). Therefore, instead of working with a continuous diameter's distribution, CoralBlox
updates each coral block separately. A mortality event is modelled as a decrease in a coral
block's density, and a growth event as an increase in a coral block's diameter range limits.
Both growth and mortality rates are size class dependent, which means the same growth and
mortality rates are used for blocks within a size class. Figure S14 shows a model's
behaviour under low disturbance and no interventions.

![Time series comparing modelled and observed relative coral cover for Ben reef showing a
recovery process under very low disturbance and no interventions. The observed data used
here is from the AIMS Long-Term Monitoring Program (LTMP). Error metrics including Root Mean
Square Error (0.06) and the Spearman Rank Correlation Coefficient (0.9) is also shown. The
second plot shows the DHW and Cyclone or COTS events. The third plot shows the benthic
composition for each functional group as a proportion of the reef’s cover at each
year.](../figs/coralblox_reef_example.png){#sfig-coralblox-example width fig-align="center"
width="14.82cm"}

In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines the proportion of the population on a reef that is
susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021] and expert opinion, and are additionally varied in the scenario
exploration.

@@

### Sensitivity analysis

A global sensitivity analysis was performed using Shapley Effects **goda etal** implemented
in the SAShE.jl package **SAShE citation**. For this analysis, 10,000 samples were simulated
for a single reef and timestep. Two metrics were considered, difference in coral cover and
difference in evenness. The factors analysed were heritability, mean colony diameter, linear
extension, mortality base rate, mean and standard deviation of the tolerance distribution,
linear extension scale factor, mortality rate scale factor reef habitable area, reef depth,
DHW, and initial coral cover. @sfig-coralblox-sa shows the result of such analysis for
different ranges of DHW. In both cases a similar effect takes place: when DHW = 0, initial
cover is the most influential factor then, when DHW \> 0, depth and DHW start to appear as
influential factors and, as DHW values get higher, depth becomes more influential in
detriment of DHW.
